---
title: "Free Agents Adapt Better than Conformist Puppets"
subtitle: "Fall Quarter Preview and Scoping Lecture"
author: "Matt Turner"
date: "2025-09-17"
categories: [learning models, formal analysis, simulation, cultural evolution]
editor: 
  markdown: 
    wrap: 80
---

# Introduction

-   Freedom to choose one's own actions based on one's self-interest is a
    hallmark of justice and equality. Human cognition lets us infer what might
    be beneficial for us to do. Human capacity for information integration
    verges on miraculous—in the blink of an eye we can identify a room full of
    people remembering rich details about them at the same time; evaluate the
    mood; and self-regulate our mood so we can react mindfully and wisely to a
    changing environment [@Witt2024]. Most of these miraculous capacities do not
    come pre-installed at birth: *homo sapiens* evolved long developmental
    childhoods and culture, both of which affect how we think, learn, and
    interact [@Tomasello2019].
-   Choosing based on what others do, i.e., conformity, might align with one's
    self-interest but it does not in general. Oftentimes conformity is
    *enforced* through coercion, including violence or threats of violence.
    -   We often communicate in "echo chambers": our social partners tend to
        have adopted similar practices as us and share similar opinions.
    -   This means that even if a sustainable practice is popular, certain
        groups may only rarely observe others doing the practice. Conformity
        would therefore be a weak or absent force for promoting sustainability
        within that group.
-   

## Sustainability

![The VIBE system helps us identify how sustainability priorities can motivate
new social science studies.](./Figures/intro-sdg-vibe.png){#fig-sdg-vibe
fig-align="center" width="950"}

### Social learning of sustainable practices

There are innumerable ways to represent the learning process. We will focus on
just two highly simplified representations that nonetheless capture the two
learning styles we want to compare:

![An individual considers whether to install solar panels, an adaptive practice.
Dollar symbols represent
"successfulness".](images/clipboard-3131304695.png){#fig-solar-example
width="600"}

-   Consider the illustration in @fig-solar-example: the *focal agent*, i.e.,
    the one we focus on who is deciding what to do.
-   There are two

### Social impediments to sustainability

-   Self-interested development groups may promote technology that they or
    business partners will benefit from installing. In reality, Indigenous
    practices often do much better.

## Representation of the cultural process

# Prevalence Dynamics

The prevalence dynamic is a way to express change in the prevalence of the
adaptive practice at each time step.

![Diagram of behavior possibilities with three learning outcomes framed in terms
of the adaptive
behavior.](images/clipboard-3172079776.png){#fig-behaviors-and-transitions
width="600"}

## Dynamical systems

### The *prevalence dynamic* equation

\newcommand{\dadt}{\frac{da}{dt}}
\newcommand{\adot}{\dot a}
\newcommand{\pal}{P(A|L)}
\newcommand{\pla}{P(L|A)}

-   $a$ is the prevalence of the adaptive behavior, $A$
-   $\frac{da}{dt}$ is the derivative of $a$ with respect to time, $t$
-   $\dot a$ is a shorthand notation for $\frac{da}{dt}$ introduced by Newton
-   $p_A$ is the total probability that a random individual, $i$, will either
    *adopt* or *keep* the adaptive behavior $A$
    -   We also use notation $p_A = P(A)$
    -   This can be expanded into the probability that an $L$-doer\
        *adopts* $A$ on the next time step, written $P(A | L)$, and the
        probability that $A$-doer keeps doing $A$ on the next time step, written
        $P(A | A)$.

$$
\frac{da}{dt}= \dot a= p_A - a
$$

## Agent-based model

```{r}
# Uncomment and install if you don't already have this.
# remotes::install_github("SocSci-for-Sustainability/socmod")

# Load other required libraries

```

Load required libraries. Use the RStudio tool to install any libraries you are
missing.

```{r}
#| output: false
library(socmod)
library(ggplot2)
library(dplyr)
```

To start we need to initialize our model and the agents using `make_abm` and
`initialize_agents` from `socmod`. We specify that there are ten agents and no
network structure by explicitly providing an empty graph (no edges) with ten
nodes to the `graph` named argument.

We set two agents to be in the seed set who receive "direct instruction", i.e.,
we set the behaviors of two randomly-chosen agents to be "Adaptive". "Adaptive"
and "Legacy" are the two default values for agent behaviors, so there is no need
to specify these. The `initial_prevalence` passed to `initialize_agents` is
implicitly the initial prevalence of the *adaptive* behavior.

```{r}
#| fig-width: 5
#| fig-height: 3.5

# Initialize the ABM and agents
abm <- 
  make_abm(graph = igraph::make_empty_graph(10)) |>
  initialize_agents(initial_prevalence = 0.2)

# Inspect the initialization to ensure 2 of 10 are A-doers
plot_network_adoption(
  abm, layout = igraph::in_circle(), 
  plot_mod = \(p) p + ggtitle("Adoption at t = 0")
)
```


### Conformity model dynamics

Below I pasted the definition of conformity learning model dynamics from
`posts/free-agents-adapt-better/R/model.R`.

```{r}
#| execute: false
#' Create a conformity learning instance of ModelDynamics.
#' 
#' Returns socmod::ModelDynamics
conformity_dynamics <- make_model_dynamics(
  
  # Partner selection is not necessary in conformity
  partner_selection = \(focal_agent, model) {},
  
  # Conformity interaction is not partner-based, so use dummy arg
  interaction = \(focal_agent, ., model) {
    
    # Observe behaviors of randomly-chosen demonstrators
    behaviors_sample <- sample(model$get_parameter("agent_behaviors"), 
                               model$get_parameter("n_demonstrators"))
    
    # Count how many of each behavior is present
    behaviors_table <- table(behaviors_sample)
    
    # Sample behavior to copy, weighted by frequency; names are behaviors
    next_behavior <- sample(names(behaviors_table), size = 1, prob = behaviors_table)
    
    # Only need to set next behavior, payoffs irrelevant w/ conformity
    focal_agent$set_next_behavior(next_behavior)
  },
  # Use the learning model stepper that makes "next" behavior/payoff "current"
  model_step = \(model) {
    learning_model_step(model)
    
    model$set_parameter("agent_behaviors", 
                        purrr::map_chr(model$agents, \(a) a$get_behavior()))
  
  }
)
```

Let's define the agent-based model again, this time with more agents. We will
also run a single trial and plot observations of the simulated prevalence
dynamics.

When I do little experiments like this I set up single cells that can run an
entire model setup and simulation trial. Then I keep running it to see what
changes.

```{r}
#| warnings: false

# Load model code w/ full project path posts/free-agents-adapt-better/R/model.R
source("R/model.R")

t1 <- single_conformity_trial(n_agents = 100, n_demonstrators = 3)
t2 <- single_conformity_trial(n_agents = 100, n_demonstrators = 3)
t3 <- single_conformity_trial(n_agents = 100, n_demonstrators = 3)
```

```{r}
#| fig-height: 6
library(patchwork)

# Plot prevalence dynamics of Adaptive behavior only
p1 <- plot_prevalence(t1, tracked_behaviors = c("Adaptive")) + ggtitle("Trial 1") + theme(plot.title = element_text(size = 12))
p2 <- plot_prevalence(t2, tracked_behaviors = c("Adaptive")) + ggtitle("Trial 2") + theme(plot.title = element_text(size = 12))
p3 <- plot_prevalence(t3, tracked_behaviors = c("Adaptive")) + ggtitle("Trial 3") + theme(plot.title = element_text(size = 12))

(p1 / p2 / p3) +
  plot_layout(guides = "collect", axes = "collect") 
```

Every time we run the series of commands above we get a new three time series of
adaptation prevalence. Some will end at a prevalence of 1.0, indicating all
agents are $A$-doers. Some end at a prevalence of 0.0, indicating all are
$L$-doers. When one behavior becomes the only behavior we say that behavior has
fixated. Typically when an intervention fails it is because of

### Success-biased model dynamics

-   We define "success" here as the fitness of one's behavior. Learners are
    biased towards teachers who are more successful in success-biased learning.
    -   This means we need to implement partner selection, which samples some
        number of demonstrators, then samples one randomly, weighted by payoff.
    -   Interaction is simple in this model: we assume perfect learning, i.e.,
        the learner always adopts its chosen teacher's behavior

```{r}

```

# References
